---
title: "Reproducible Research - Project I"
author: "Dalip Sondhi"
date: "Sunday, June 07, 2015"
output: html_document
---

Start the timer to monitor procesing time.

```{r}
start <- proc.time()
```

### Introduction

This analysis uses the [activity monitoring](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) dataset for analysis of the steps measurements taken over a 61-day period. The steps were measured at 5-minute intervals.


Load the packages that will be used in this analysis.

```{r}
suppressWarnings(suppressMessages(library(dplyr))) 
suppressWarnings(suppressMessages(library(lubridate)))
suppressWarnings(suppressMessages(library(lattice)))
```

Read the dataset into the variable *activity* and check its structure:

```{r}
activity <- read.csv("activity.csv")
str(activity)
```

Change the date to date format.

```{r}
activity$date <- ymd(as.character(activity$date))
```

Check a few rows of the dataset.

```{r}
head(activity)
```

Find the dimensions and unique dates for the dataset.

```{r}
dim(activity)
uDates <- unique(activity$date)
length(uDates)
```

### Part I

Create a subset *activityf* with NAs filtered out.

```{r}
activityf <- filter(activity, steps>=0)   #remove NAs
```

Find the number of steps taken daily.

```{r}
bydate <- group_by(tbl_df(activityf), date)
daily <- summarise(bydate, Steps=sum(steps))
dim(daily)
```

Note the histogram below has 53 days excluding the 8 days where all steps were NAs:

```{r Figure_1_Histogram, fig.width=5, fig.height=4}
hist(daily$Steps, 
     xlab   = "Number of Daily Steps", 
     main   = "Histogram of Daily Steps", 
     ylim   = c(0, 30), col="skyblue", 
     labels = TRUE)
```

Find the mean and median of the number of daily steps:

```{r}
mean(daily$Steps)
median(daily$Steps)
```

For the time-series plot, this was assumed to require an average of each 5-min interval across all days:

```{r Figure_2_Time_Series_Plot}
byinterval <- group_by(tbl_df(activityf), interval)
fiveMin <- summarise(byinterval, Steps=sum(steps), count=n())
fiveMin <- mutate(fiveMin, meanSteps=Steps/count)

plot(fiveMin$interval, fiveMin$meanSteps, 
     main="Time Series Plot of 5-min Interval Means",
     xlab="Interval Number",  
     ylab="Mean Steps",
     col="blue",
     type="l")

abline(h=mean(fiveMin$meanSteps), lty=2, lwd=1.5, col="red")  #show the mean of interval means
text(260, 48, "mean=37.4", cex=0.8, col="red")
```

Find which interval contains the maximum steps.

```{r}
maxinterval <- fiveMin[which(fiveMin$meanSteps==max(fiveMin$meanSteps)),]
maxinterval
```

The interval number 835 has the maximum mean steps of 206.17.


### Part II

Find the number of rows with steps showing NAs in the *activity* dataset.

```{r}
nulls <- is.na(activity$steps)    #This is the logical vector showing TRUE for NA and FALSE otherwise
```

Find the number of NAs. 

```{r}
length(nulls[nulls==TRUE])       #This provides the number of NAs
```

Create a new dataset *activityNew* that replaces the NAs. Replace the NAs by the 5-min average based on the mean od the daily steps excluding NAs, i.e. 10766.19 shown earlier. Convert this to a 5-min value. There are 24 hr x 12 5-min intervals/hr = 288 5-min intervals per day. So there are 10766.19/288 = 37.3826 average steps in each 5-min interval. Replace the 2304 NAs with this average value.

```{r}
activityNew <- activity
activityNew$steps[nulls] <- mean(daily$Steps)/288
activityNew$date <- ymd(as.character(activityNew$date))  # change to date format
dim(activityNew)        # check dimensions to ensure full dataset
head(activityNew)       # check a few rows
```

Find the number of steps taken daily.

```{r}
bydate <- group_by(tbl_df(activityNew), date)
daily <- summarise(bydate, Steps=sum(steps))
dim(daily)
```

Note the histogram below has 61 days including the 8 days where all steps with NAs were replaced with the mean value of all intervals:

```{r Figure_3_Histogram_2, fig.width=5, fig.height=4}
hist(daily$Steps, 
     xlab   = "Number of Daily Steps (NAs replaced)", 
     main   = "Histogram of Daily Steps", 
     ylim   = c(0, 40), col="green", 
     labels = TRUE)
```

Display the mean and the median of the daily steps in the *activityNew* dataset.

```{r}
mean(daily$Steps)
median(daily$Steps)
```

Note that the mean is identical to the mean of the original dataset mainly because of the strategy used to replace the NAs. This need not be true if a different method had been used to replace the NAs. The median changed slightly, from 10765 to 10766.19. 

Next, create a variable with two levels - weekday and weekend in the dataset *activityNew*.

```{r}
for (i in 1:nrow(activityNew)) 
{
if (weekdays(activityNew$date[i]) == "Saturday" | weekdays(activityNew$date[i]) == "Sunday") 
     {activityNew$day[i] <- "weekend"}
else {activityNew$day[i] <- "weekday"}
}
activityNew$day <- as.factor(activityNew$day)
str(activityNew)
```

Summarize the steps by interval and day type, and plot the results

```{r Figure_4_Time_Series_Plot_2, fig.width=6, fig.height=7}
bydayinterval <- group_by(tbl_df(activityNew), day, interval)
fiveMin <- summarise(bydayinterval, Steps=sum(steps), count=n())
fiveMin <- mutate(fiveMin, meanSteps=Steps/count)

xyplot(meanSteps ~ interval | day, data=fiveMin, 
       type   = "l",
       as.table = TRUE,
       xlab   = "Interval Number",
       ylab   = "Mean Steps")
```

Beyond this point, not asked for but compare the means for weekdays and weekends - they appear to be different

```{r}
gp1 <- group_by(fiveMin, day)
summarise(gp1, meanSteps=mean(meanSteps))
```
Run the *t.test* to determine if the means are significantly different.

```{r}
wk <- filter(fiveMin, day=="weekday")
we <- filter(fiveMin, day=="weekend")
t.test(wk$meanSteps, we$meanSteps, paired=FALSE)
```
Yes, the means are higher during weekends as might be expected (*p-value < 0.05*)!

Check the timer.

```{r}

time <- proc.time() - start
time
```